Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2014-06-25 improved_state_space.tex
Response to reviewer's comments

\documentclass[12pt,english]{article}
%\renewcommand*\abstractname{Summary}
\usepackage[verbose,letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} %,displaymath
\linenumbers
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage{indentfirst}
\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
%\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage[pdftex,hidelinks,pdfstartview={XYZ 1 1 1.5}]{hyperref}
\usepackage{rotating}
\usepackage[compact]{titlesec}
\titleformat{\section}[display]{\normalfont\bfseries}{}{0em}{\centering}
\titleformat{\subsection}[display]{\normalfont\itshape}{}{0em}{\centering}
\bibliographystyle{ecology}
\setlength{\bibsep}{0.0pt}
\bibpunct{(}{)}{,}{a}{}{,}
%\hyphenpenalty=100000
\setlength{\jot}{-1.2ex}
\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\doublespacing}

\begin{document}
\renewcommand{\thefootnote}{\arabic{footnote}}
\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data}

%\doublespacing
\author{Jeffrey A. Hostetler\thanks{Migratory Bird Center, Smithsonian 
Conservation Biology Institute, National Zoological Park, MRC 5503, 
Washington, DC 20013-7012}
   \and Richard B. Chandler\thanks{USGS Patuxent Wildlife Research Center 
        Laurel, MD 20708} \footnote{Current address: Warnell School of
     Forestry and Natural Resources, University of Georgia,
     \url{rchandler@warnell.uga.edu}, +001-706-542-5815}        
}
\date{} %remove date
\maketitle
\begin{spacing}{1.9}
\begin{flushleft}
\abstract{%\noindent 
Models of population dynamics %play a central role in theoretical and
%applied ecology, where they 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by observation error, which
can inflate estimates of process variation if ignored.  
%Failure to account for observation error in such models can result in 
%bias. Therefore, ecologists have increasingly relied on 
Recently, state-space models have been developed to deal with this problem
%these probdirectly model 
by directly modeling both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
several important limitations: (1) they assume that random effects are
Gaussian distributed, which implies that abundance can be negative and
that false positive observation errors are equally likely as false negative
errors; %results in unrealistic predictions
%(e.g. negative values of abundance) and an implausible mode
%and observation error are Gaussian random effects
%the parameters are not identifiable
%in many common situations, 
(2) they do not admit spatial variation in
population dynamics; and (3) some of the parameters of the model are not
estimable. % they assert an implausible model for
%we have no clear interpretation of the
%observation error. 
We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} 
%for spatially-replicated time series data 
that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects. 
% such as observer-specific detection probabilities. 
We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. %We provide code to fit these models using the \textbf{R} package
%\texttt{unmarked} and \textbf{JAGS} is also provided.
Results of a simulation study suggest that bias is %was 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
}

@@

%\newpage
\vspace{0.5cm}

\section*{Introduction}
%Theoretical ecology requires models of population dynamics for testing
%hypotheses regarding spatial and temporal variation in
%abundance. For example, much theoretical work has focused on
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, %the have been used to assess
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999,royama:1977}. %,dennis_taper:1994
%and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

Several challenges are routinely encountered when trying to fit models
of population dynamics to data from field studies. % is that 
%In order to address these questions, two complicating factors must
%be confronted when fitting population models to data. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

Recently developed state-space models %are a widely used approach for
have made it possible to %become the primary tool for 
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
%process variation and 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be overestimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
% Richard, is switching this from $\grave{o}$ to $\eta$ okay?  Andy thought
% the o with accent was weird and I think I agree.  $\nu$ seems standard,
% but we use it for a slightly different meaning (ES only) later.  
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
A generic model can be written: %The full model can now be written as:
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \nonumber \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
%where $\eta_t$ is the random effect allowing for process
%variation unaccounted for by the deterministic model. 
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. %Standard practice is to
%ignore 
The process variation associated with $N_1$ is often ignored, or
%, as indicated by Eq~\ref{eq:ss1a}. Alternatively, 
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
overcome many of the limitations associated with the application of
classical models of population dynamics, 
%are among the most widespread approaches for modeling population dynamics
%using time series data, 
several problems exist. %, which we explain in the following paragraphs. 
% First,
% time series data and the underlying abundance parameters of interest
% are typically integer-valued, as in the case of count data, raising
% concerns about the Gaussian assumption for the random
% effects. Second, the model for observation error is implausible. %has little biological basis. 
% Third, the model does not account for spatial variation in abundance. Lastly, some of the
% parameters of the model are not estimable. We briefly discuss each of
% these points before describing a general approach to resolving these
% shortcomings.
%Use of the Gaussian distribution for modeling random process variation
%and observation error is often motivated by convenience rather than
%biology. Specifically, t
The assumption that random effects are Gaussian distributed makes it
easier to estimate parameters because methods such as the Kalman
filter can be applied %The Gaussian assumption allows for efficient methods of parameter
%estimation, such as the Kalman filter
\citep{dennis_etal:2006}; however, % which are not available otherwise % which is much more computationally efficient
%than estimation methods when random effects are not Gaussian distributed
%\citep{devalpine_hastings:2002}. %The problem with this is that it
%However, 
these models effectively assume that abundance can be take on any
real value, and hence they can predict negative values of abundance. %including negative values, 
%This approach allows for negative and non-integer values of the state-variable, which is
%inconsistent with the observed data, and may 
%and these models can result in implausible predictions.
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

With respect to the observation error, the Gaussian assumption
%Another problem with standard state-space models is that the model for
%the mean zero Gaussian distribution for the observation error
implies that false positive detections are
equally likely as false negatives, a pattern inconsistent with most
findings \citep{miller_etal:2011}. 
%. To our knowledge, all studies to
%date have shown that false negative detections are much more common
%that false positives \citep{miller_etal:2011}.
%will be
%higher than $N_t$ as often as it is lower than $N_t$. 
%Identifying a biological mechanism that would cause such symmetric errors
%is difficult.
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as its elusiveness,
or to individuals being missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models of population dynamics \citep[but
see][]{buckland_etal:2004}.

Other limitations of state-space models, as commonly applied in ecology,
include the failure to admit spatial variation in abundance and
the fact that some parameters are not identifiable
\citep{polansky_etal:2009}. 
% without making 
%Another limitation of many state-space models is that they do 
%Common formulations of state space models do 
%not admit spatial variation in abundance or demographic
%parameters. 
Ignorning spatial variation simply reduces the scope of the
inferences, 
% that can 
%be drawn from the models, it disregards the fact that space is an
%important factor in the regulation of natural populations \citep{murdoch:1994,
%pulliam:1988, gill_etal:2001}.
but identifiability problems require %lack of identifiability because
model constraints or the assertion of additional assumptions. 
%A more serious problem %than the ones associated with the Gaussian assumptions 
%is that the parameters of simple state-space models
%such as Eq~\ref{eq:ss1} are not identifiable in many
%circumstances \citep{polansky_etal:2009}. 
Specifically, abundance in
the first year cannot be estimated because there is only one data
point (the count in the first year) that carries any information about
it. To remedy this, pratitioners either assume that there is no
observation error in the first year, or they assume the population is
at equilibrium, which may defeat the purpose of many studies of
population dynamics. 
%Recall that the
%Eq.~\ref{eq:ss1} did not specify distributions for $N_1$, % or $X_1$;
%although $N_1$ is a random variable. Thus, uncertainty exists
%that should be accounted for in the model.  
%%are needed to fully specify the model.
%To be more specific, a fully-specified state-space model requires at least
%three probability distributions, which we represent
%using bracket notation (i.e., $[Y|\Omega]$ is interpreted
%as the probability distribution of the random variable $Y$ given the
%parameter $\Omega$). The three probability distributions required for a
%state-space model are:
%%%% JEFF, does this extra sentence help? This notation comes from
%%%% several of the original state-space papers and is widespread in
%%%% books like Andy's and Bill's.
%\hl{[I'm not familiar with bracket notation,
% so these equations are confusing to me. How widely used is this
% notation in ecology? Will others likely understand this? Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
%\begin{subequations}
%  \label{eq:ss2}
%  \begin{align}
%    [N_1&|\bm{\theta}] \label{eq:ss2a} \\ \label{eq:ss2b}
%[N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\
%\label{eq:ss2c}
%    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
%  \end{align}
%\end{subequations}
%where $\bm{\theta}$ are the process variation parameters for the
%initial state, $\bm{\Theta}$ are the process variation parameters
%influencing how abundance changes over time, and $\bm{p}$ are the
%observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
%and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
%models, but what should be the distribution for Eq.~\ref{eq:ss2a}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}. However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%distribution, $[N_1|\bm{\theta}]$ be?
%And how can $\bm{\theta}$ be estimated given that there is only a single observation
%vailable? 
%Often, these problems are simply assumed away. For
%example, sometimes researchers assume that there is no process variation in
%the first year, although this seems unjustified. Alternatively, the
%population may be assumed to be at equilibrium such that the expected
%value of $N_t$ is constant through time. Although this 
%makes the parameters $\bm{\theta}$ identifiable, assuming equilibrium
%defeats the objective of many studies of population dynamics, namely
%determining why a population varies over time.
%% OR
%Sometimes researchers assume that there is no process variation in 
%the first year or that the population is at equilibrium such that the expected
%value of $N_t$ is constant through time, but these approaches are unsatisfactory.

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. Observation models with
more intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of population models. %This
%limits the utility of their model for making inferences about explicit
%population processes.

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves each of the problems outlined above. 
%simultaneous resolves each of the aforementioned problems with
%traditional state-space models, and is designed for simple count data.
%In the following section, we describe the DM model in its original
%form and explain how it resolves each of the deficiencies with
%standard state-space models. 
%In Section~\ref{sec:ext}, 
Our aim is to extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time series, such as excess zeros and unexplained random variation. 
%Specifically, we describe methods for accommodating
%excess zeros and nuisance variables such as random observer effects.
Both frequentist and Bayesian methods of inference are discussed, and
we evaluate the performance of the model using a simulation study and by
analyzing data from the North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}.
%code for fitting models is presented in the appendices.
%The Bayesian method
%is attractive in that it can accommodate prior information about
%detection parameters, which is available for many species, yet not
%collected as part of many existing monitoring programs.
%In Section~\ref{sec:app}, we evaluate the performance of the model
%extensions using a simulation study and by analyzing data from the
%North American Breeding Bird Survey (BBS), one of
%the most spatially and temporally extensive sets of count data on
%vertebrate populations \citep{robbins_etal:1986}. %The overarching aim
%of the paper is to provide ecologists with flexible and accessible means of
%addressing important questions related to the variation of abundance in
%space and time. 
%believe these extensions will increase the utility of the model for
%application to existing datasets such the Breeding Bird Survey.

@@

%\subsection{The Data}

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. 
%A site is ideally a
%well-defined region of a study area, such as a wetland or a
%patch of early-successional habitat, although it could be an arbitrarily
%defined region such as a randomly located survey plots. The time
%frame of the study is arbitrary, but in general
%it will be sufficiently long to allow for temporal variation in
%abundance. For instance, a primary sampling period could be a
%one month long breeding season, which could be surveyed once per
%year for $T$ years. If the a site is surveyed on $J$ occasions during a
%primary period, these are called secondary sampling occasions.
%In the case where no secondary sampling was conducted, i.e. $J=1$, 
Let $X_{i,t}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. 
%If secondary sampling periods were
%used, an extra dimension is added so that we have $X_{ijt}:
%j=1,\hdots,J$. 
In general, $X_{i,t}$ will be lower than abundance, 
%actual quantity of interest, abundance,
%denoted 
$N_{i,t}$, but in cases where detection is perfect,
abundance is observed directly such that $X_{i,t} \equiv N_{i,t}$.

%\subsection{The Original Formulation of the Model}

Like traditional state-space models, the DM model includes 
three conditionally related processes corresponding to: 
% shown in Eq.~\ref{eq:ss1}, which correspond to 
(1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process %simply 
describes the the relationship between
abundance and the observed count data.

@@

%Recall that 
Conventional state-space models assume that the
distribution for the initial time period %,
%$[N_1 | \bm{\theta}]$, 
is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i1}$
as either a Poisson or negative binomial random variable:
\begin{linenomath*}
\begin{gather}
N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
\end{linenomath*}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.
%In the Poisson case, the equivalent expression of
%$[N_1 | \bm{\theta}]$ (Eq.~\ref{eq:ss2a}) is
%$[N_{i1} | \Lambda] =\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}$.
%\begin{equation}
%[N_{i1} | \Lambda] =
%\frac{\Lambda^{N_{i1}}e^{-\Lambda}}{N_{i1}!}.
%  \label{eq:N1br}
%\end{equation}
% Maybe we don't need the Poisson distriubtion equation? - Jeff

Regardless of the specified distribution, the model for initial
abundance has two distinguishing features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
\beta^{\Lambda}_1
x_{i}$. The second important point is that the spatial
replicates resolve the
problem of parameter non-identifiability that are common to
standard state-space models because, as demonstrated by
\citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data. Hence, the first component of
the model addresses both the issues of spatial inference and
parameter identifiability discussed previously.

@@

%Eqs.~\ref{eq:N1} and \ref{eq:Nt} fully specify the state model,
%i.e. the model for abundance dynamics. This state
%model has allowed us to replace the initial abundance
%distribution and transition models of conventional state-space models, with
%alternatives that (1) allow for modeling spatial variation in
%abundance, (2) respect the discrete nature of the counts, and
%(3) have parameters that are estimable.
%The one
%remaining task is to replace the observation model
%$X_t = N_t + \zeta_t$ (Eq.~\ref{eq:ss1c}) with one that can be
%more easily interpreted. 
The observation model adopted by \citet{dail_madsen:2011} is the same
binomial model proposed by \citet{royle:2004biom}: % assumes that individuals are missed due to imperfect
%detection. The simplest model for imperfect detection is
\begin{linenomath*}
\begin{equation}
  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
  \label{eq:p1}
\end{equation}
\end{linenomath*}
where $p$ is the probability of detecting each individual. Variation
in detection probability can be modeled as a function of site-specific
or occasion-specific covariates using, for example, a logit-linear model.
%or equivalently,
%\begin{equation}
%[X_{it} | N_{it}, \bm{p}] = \binom{N_{it}}{X_{it}}
%\bm{p}^{X_{it}}(1-\bm{p})^{N_{it}-X_{it}}.
%  \label{eq:1}
%\end{equation}
%Under this model, $\bm{p}$ is interpreted as the detection
%probability parameter.

>>>>

\documentclass[12pt,english]{article}
\usepackage[verbose,letterpaper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines]{lineno} 
\linenumbers
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,comma,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage{indentfirst}
\usepackage[format=plain,justification=raggedright,labelsep=period,font={stretch=1.9}]{caption}
\usepackage[pdftex,hidelinks]{hyperref} 
\usepackage{rotating}
\usepackage[compact]{titlesec}
\titleformat{\section}[display]{\normalfont\bfseries}{}{0em}{\centering}
\titleformat{\subsection}[display]{\normalfont\itshape}{}{0em}{\centering}
\bibliographystyle{ecology}
\setlength{\bibsep}{0.0pt}
\bibpunct{(}{)}{,}{a}{}{,}
\setlength{\jot}{-1.2ex}
\usepackage{etoolbox}
\AtBeginEnvironment{tabular}{\doublespacing}
\makeatletter
\let\@fnsymbol\@arabic
\makeatother
\title{Improved state-space models for inference about 
spatial and temporal variation in abundance from count data}

\author{Jeffrey A. Hostetler\thanks{Migratory Bird Center, Smithsonian 
Conservation Biology Institute, National Zoological Park, MRC 5503, 
Washington, DC 20013-7012} \textsuperscript{,}\footnote{Email: \href{mailto:hostetlerj@si.edu}{hostetlerj@si.edu}}
   \and Richard B. Chandler\footnote{USGS Patuxent Wildlife Research Center 
        Laurel, MD 20708} \textsuperscript{,}\thanks{Current address: Warnell School of
     Forestry and Natural Resources, University of Georgia}        
}
\date{} %remove date
\begin{document}
\maketitle
\vspace{-1cm}
\begin{spacing}{1.9}
\begin{flushleft}
\abstract{
Models of population dynamics 
are frequently used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Fitting models
of population dynamics to field data is challenging because most data
sets are characterized by observation error, which
can inflate estimates of process variation if ignored.  
Recently, state-space models have been developed to deal with this problem
by directly modeling both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
several important limitations: (1) they assume that random effects are
Gaussian distributed, which implies that abundance can be negative and
that false positive observation errors are equally likely as false negative
errors; (2) they do not admit spatial variation in
population dynamics; and (3) some of the parameters of the model are not
estimable. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} 
that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects. 
We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, as demonstrated by accompanying {\bf R} and {\bf JAGS} code. 
Results of a simulation study suggest that bias is 
negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlights how these methods
can be readily applied to existing data, but it also suggests 
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is lacking.
}

@@

\section*{Introduction}
Models of population dynamics are vital in both theoretical and
applied ecological research. For instance, 
the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and
spatial synchrony have been studied by comparing data from natural
populations with theoretical models
\citep{may:1975,turchin:1990,bjornstad_etal:1999}. 
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.

Several challenges are routinely encountered when trying to fit models
of population dynamics to data from field studies. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

Recently developed state-space models 
have made it possible to
study population dynamics while accounting for both process variation
and observation error \citep{devalpine_hastings:2002,
  buckland_etal:2004, dennis_etal:2006}. Classical state-space
models are time series models in which the true state of the
system (e.g. population size during each year) is modeled as a hidden
process, and the observed data are modeled conditional on this process
and the observation error. One reason for the widespread adoption of
state-space models in ecology is that failure to account for 
observation error can bias estimators of abundance and
population growth parameters. For instance, the strength of
density dependence will be overestimated if observation error is
ignored \citep{link_nichols:1994,shenk_etal:1998}.

A simple state-space model can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\eta_t$.
The growth model may be density-dependent, as in the case of the 
Ricker-logistic model, or it might be density-independent, such as when growth
is exponential.
A generic model can be written: 
\begin{linenomath*}
\begin{gather}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \nonumber \\
N_t &= \mu(N_{t-1}) + \eta_{t-1} \quad \text{for} \;
t=2,\hdots,T  \\
X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \;
t=1,\hdots,T. \nonumber 
  \end{align}
\end{gather}
\end{linenomath*}
In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\eta_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. 
The process variation associated with $N_1$ is often ignored, or
the population is assumed to be at equilibrium such that the $N_1$ can
be regarded as an outcome of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
overcome many of the limitations associated with the application of
classical models of population dynamics, 
several problems exist. 
The assumption that random effects are Gaussian distributed makes it
easier to estimate parameters because methods such as the Kalman
filter can be applied 
\citep{dennis_etal:2006}; however, 
these models effectively assume that abundance can be take on any
real value, and hence they can predict negative values of abundance. 
Transforming the count data and the underlying abundance variables 
do little to resolve these issues \citep{ohara_kotze:2010}.

With respect to the observation error, the Gaussian assumption
implies that false positive detections are
equally likely as false negatives, a pattern inconsistent with most
findings \citep{miller_etal:2011}. 
A more likely form of observation error, and one that has been recognized for well
over a century, results from failing to detect individuals that are
present. Imperfect detection may be attributable to
characteristics of the species under study, such as it being present but unavailable for detection,
or to individuals being available but missed by the ecologist collecting the data in the field.
Although a vast number of methods have been devised for accounting for
this form of observation error, rarely have these methods been
integrated into state-space models of population dynamics \citep[but
see][]{buckland_etal:2004}.

Other limitations of state-space models, as commonly applied in ecology,
include the failure to admit spatial variation in abundance and
the fact that some parameters are not identifiable
\citep{polansky_etal:2009}. 
Ignoring spatial variation reduces the scope of the
inferences and can lead to underestimation of variances. 
Identifiability problems require 
model constraints or the assertion of additional assumptions. 
Specifically, abundance in
the first year cannot be estimated because there is only one data
point (the count in the first year) that carries any information about
it. To remedy this, practitioners either assume that there is no
observation error in the first year, or they assume the population is
at equilibrium, which may defeat the purpose of many studies of
population dynamics.

Several extensions of state-space models have been proposed to
overcome the limitations described above. \citet{devalpine_hastings:2002} and
\citet{brooks_etal:2004} described methods for fitting models with non-Gaussian
distributions for the process and observation errors. Observation models with
more intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by 
\citet{kery_etal:2009}. \citet{lele_etal:1998} and 
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, only
the work by \citet{kery_etal:2009} addressed several of these limitations
simultaneously; however, their model did not include serial
dependence, which is a hallmark of population models.

In this paper, we focus on the model of \citet[henceforth the DM model]{dail_madsen:2011}
that resolves each of the problems outlined above. 
Our aim is to extend the
model to accommodate classical models of population growth and
to handle several features common to ecological
time series, such as excess zeros and unexplained random variation. 
Both frequentist and Bayesian methods of inference are discussed, and
we evaluate the performance of the model using a simulation study and by
analyzing data from the North American Breeding Bird Survey (BBS), one of
the most spatially and temporally extensive sets of count data on
vertebrate populations \citep{robbins_etal:1986}.

@@

The DM model requires count data collected at $R$ sites, each of
which is surveyed on $T$ primary sampling periods. 
Let $X_{i,t}: i=1,\hdots,R; t=1,\hdots,T$ denote the count data
at site $i$ and primary period $t$. 
In general, $X_{i,t}$ will be lower than abundance, 
$N_{i,t}$, but in cases where detection is perfect,
abundance is observed directly such that $X_{i,t} \equiv N_{i,t}$.

Like traditional state-space models, the DM model includes 
three conditionally related processes corresponding to: 
(1) initial abundance, i.e. the
abundance at site $i$ during the first primary period,
denoted $N_{i,1}$, (2) abundance at time $t$ (for $t>1$) which depends upon
abundance at $t-1$, and (3) the
detection process \citep{dail_madsen:2011}.
The first two processes describe the state process---the
variation in abundance in space and time. The third process 
describes the the relationship between
abundance and the observed count data.

@@

Conventional state-space models assume that the
distribution for the initial time period 
is either the equilibrium
distribution, or has zero variance.
In contrast, \citet{dail_madsen:2011} proposed modeling $N_{i,1}$
as either a Poisson or negative binomial random variable:
\begin{linenomath*}
\begin{gather}
N_{i,1} \sim \mathrm{Pois}(\Lambda) \nonumber \\
\text{or} \nonumber \\
N_{i,1} \sim \mathrm{NB}(\Lambda, \alpha)
\label{eq:N1}
\end{gather}
\end{linenomath*}
where $\Lambda_i$ is the expected abundance at site $i$ during
year 1.
The Poisson distribution assumes that the mean of $N_{i,1}$ is
equal to its variance, whereas the negative binomial distribution allows the
variance to be greater than the mean with the amount of
overdispersion determined by the parameter $\alpha$.

Regardless of the specified distribution, the model for initial
abundance has two distinguishing features. First, it provides a
mechanism for characterizing spatial variation in abundance. For
instance, one might consider the influence of some environmental
covariate ($x_i$) on abundance using a log-linear
model such as $\log(\Lambda_i) = \beta^{\Lambda}_0 +
\beta^{\Lambda}_1
x_{i}$. Second, the spatial
replicates resolve the
problem of parameter non-identifiability that are common to
standard state-space models because
%, as demonstrated by \citet{royle:2004biom},
process variation and observation error can be estimated from
spatially-replicated count data \citep{royle:2004biom}. Hence, the first component of
the model addresses both the issues of spatial inference and
parameter identifiability discussed previously.

@@

The observation model adopted by \citet{dail_madsen:2011} is the same
binomial model proposed by \citet{royle:2004biom}: 
\begin{linenomath*}
\begin{equation}
  X_{i,t} \sim \mathrm{Bin}(N_{i,t}, p)
  \label{eq:p1}
\end{equation}
\end{linenomath*}
where $p$ is the probability of detecting each individual. Variation
in detection probability can be modeled as a function of site-specific
or occasion-specific covariates using, for example, a logit-linear model.

\section*{Model Extensions}
\label{sec:ext}
\subsection*{Population growth models}
\begin{samepage}
Partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, but this is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. 
When the mechanistic model is unrealistic, we suggest replacing it
with classical population growth models that have a long history in ecology.
The exponential growth model is the simplest: 
$N_{i,t} = N_{i,t-1}e^r$ where $r$ is the intrinsic
rate of increase. 
To allow for demographic stochasticity, we can regard $N_{i,t}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{linenomath*}
\begin{equation}
  N_{i,t} \sim \mathrm{Pois}(\exp(r)N_{i,t-1}).
\label{eq:exp}
\end{equation}
\end{linenomath*}
